Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-06-04 paper2.qmd
section:6; process:write

Drawing conclusions about the real world from observed data always requires assumptions, regardless of whether the data are observational or experimental [@Kohler_et_al_2019; @Deffner_et_al_2022]. This study is no exception to this principle. By adopting causal inference as the theoretical backbone of our proposed framework, we necessarily rely on expert knowledge and assumptions about the variables' causal structure that are often untestable at the outset [@Hernan_et_al_2025]. As such, our framework does not yield automatic answers. Instead, it encourages researchers to formulate precise questions and to explicitly articulate their assumptions to develop a generalizable understanding of the CJ system [@Rohrer_et_al_2022; @Deffner_et_al_2022; @Sterner_et_al_2024]. This clarity is essential because a model's ability to produce accurate estimates and valid inferences depends heavily on how well the data and inferential goals align with its underlying assumptions [@Kohler_et_al_2019]--a property we cannot yet guarantee for our models. However, this theory-driven approach offers a key benefit: its lays the foundation for empirically testing the proposed causal assumptions [@Deffner_et_al_2022].

@@

<!-- 1. Challenges for the framework: background knowledge and research context -->
In light of these limitation, this study presents several challenges that researchers must navigate--challenges common to many quantitative studies. These challenges fall into two main categories. The first concerns the foundational knowledge researchers need to apply our proposed framework effectively. The second involves the increased attention researchers must now devote to their specific research context.

@@

In terms of foundational knowledge, researchers must face two core challenges to use our framework effectively. First, they need a solid understanding of causal inference principles. Second, they must learn to translate the functional and probabilistic content of conceptual models into bespoke statistical models. One example that illustrates the importance of a sound knowledge of causal inference is the recurrent assumption that predictor variables are "relevant" to the research context—interpreting this relevance as their inclusion in a *sufficient adjustment set* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014]. However, we do not fully explore what this assumption entails or its implications for model specification, estimation, and inference. As a result, researchers must engage deeply with these and other complex ideas, such as how SCMs encode functional and probabilistic information. To assist this effort, we include key references throughout the study to guide motivated readers toward deeper understanding[^10].

@@

Moreover, developing the skills to translate functional and probabilistic information into bespoke statistical models has its challenges. Although we argue that Bayesian inference methods offer a more accessible path for many applied researchers to develop these skills--since they reduce the need for specialized knowledge in areas like optimization theory, which frequentist methods often require--these methods still demand familiarity with technical concepts, such as probabilistic programming languages (PPLs), probability distributions, and convergence, that researchers must acquire. Nevertheless, to support researchers in building these skills, we include links to model code and alternative specifications in the **Declarations** section of this document.

@@

Regarding attention to the research context, researchers face two additional challenges. First, they must verify that the conceptual model provides a *faithful* [^11] representation of the CJ system under study. Second, they must assess whether the statistical translation of the model can accurately estimate the intended estimands (i.e., parameters) from empirical data. To ensure a faithful representation of the CJ system, we encourage researchers to treat our conceptual and statistical models as starting points rather than universal solutions for all CJ designs or datasets. As such, researchers must tackle the challenge of adapting these models to their specific research contexts, paying close attention to causal assumptions, experimental design, and model specification. Nonetheless, our approach offers a transparent framework that enables researchers to explicitly articulate any new assumptions and design CJ experiments that support reliable and valid trait measurement.

[^11]: Avoid confusing this term with the *faithfulness assumption* described in the causal inference literature. For further details, see footnote 1.

@@

Although performing identification analysis through formal derivations may appear to be a natural next step, we argue that the complexity of the CJ system renders such an approach impractical at the outset. Instead, simulation-based methods, such as power analysis, provide a more practical and flexible alternative, enabling researchers to examine the consistency of estimates without relying on cumbersome mathematical proofs. Nonetheless, the development of formal identification proofs remains a significant goal for future CJ research. Notably, our framework supports both approaches, providing a foundation for formal derivations as well as simulation-based methods.

>>>>

Drawing conclusions about the real world from observed data always requires assumptions, regardless of whether the data are observational or experimental [@Kohler_et_al_2019; @Deffner_et_al_2022]. This study is no exception to this principle. By adopting causal inference as the theoretical backbone of our proposed framework, we necessarily rely on expert knowledge and assumptions about the variables' causal structure that are often untestable at the outset [@Hernan_et_al_2025]. As such, our framework does not yield automatic answers. Instead, it encourages researchers to formulate precise questions and to explicitly articulate their assumptions to develop a generalizable understanding of the CJ system [@Rohrer_et_al_2022; @Deffner_et_al_2022; @Sterner_et_al_2024]. This clarity is essential because a model's ability to produce accurate estimates and valid inferences depends heavily on how well the data and inferential goals align with its underlying assumptions [@Kohler_et_al_2019]--a condition we cannot yet guarantee for our proposed models. Nevertheless, a key strength of this theory-driven approach is that it lays the foundation for empirically testing the proposed causal assumptions [@Deffner_et_al_2022], as discussed in @sec-discussion_RA.

@@

<!-- 1. Challenges for the framework: background knowledge and research context -->
This theoretical commitment to causal inference also introduces several challenges that researchers must navigate--challenges common to many quantitative studies. These fall into two main categories. The first concerns the foundational knowledge researchers need to apply our proposed framework effectively. The second involves the increased attention researchers must now devote to their specific research context.

@@

To effectively apply our framework, researchers must tackle two main challenges related to foundational knowledge. First, they need a solid understanding of causal inference principles. Second, they must learn to translate the functional and probabilistic content of conceptual models into bespoke statistical models. One example that illustrates the importance of a sound knowledge of causal inference is the recurrent assumption that predictor variables are "relevant" to the research context--interpreting this relevance as their inclusion in a *sufficient adjustment set* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014]. However, we do not fully explore what this assumption entails or its implications for model specification, estimation, and inference. Consequently, researchers must deeply engage with these and other complex ideas, including how SCMs encode functional and probabilistic information. To assist this effort, we include key references throughout the study to guide motivated readers toward a deeper understanding of these concepts [^10].

@@

Developing the skills to translate conceptual models into bespoke statistical models also presents challenges. Although Bayesian inference methods offer a more accessible path for many applied researchers to develop these skills--by reducing the need for specialized knowledge in areas like optimization theory, which frequentist methods often require--they still demand familiarity with other technical concepts such as probabilistic programming languages (PPLs), probability distributions, and convergence, which researchers may find hard to master. Thus, to support this skill development, we include links to the statistical model code and alternative specifications in the **Declarations** section of this document.

@@

Researchers face two additional challenges when applying our framework to their specific research context. First, they must verify that the conceptual model provides a *faithful* [^11] representation of the CJ system under study. Second, they need to assess whether the statistical translation of the model can accurately estimate the intended estimands (i.e., parameters) from empirical data. To ensure a faithful representation of the CJ system, we encourage researchers to treat our conceptual and statistical models as starting points rather than universal solutions for all CJ designs or datasets. This consideration means researchers must tackle the challenge of adapting these models to their specific research contexts, paying close attention to causal assumptions, experimental design, and model specification. Our approach supports this process by offering a transparent framework that helps researchers explicitly articulate these new assumptions and supports the design of CJ experiments that can yield reliable and valid inferences.

[^11]: Avoid confusing this term with the *faithfulness assumption* [@Neal_2020; @Hernan_et_al_2025] described in the causal inference literature. For further details, see footnote 1.

@@

Although performing identification analysis through formal derivations may appear to be a natural next step, we argue that the complexity of the CJ system renders such an approach impractical at the outset. Instead, simulation-based methods, such as power analysis, provide a more practical and flexible alternative, enabling researchers to examine the consistency of estimates without relying on cumbersome mathematical proofs. Nonetheless, the development of formal identification proofs remains a significant goal for future CJ research. Importantly, our framework supports both approaches, as it provides the probabilistic foundation for formal derivations and the necessary statistical structure for simulation-based methods.
